Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-11-16 paper2_JNLS/paper.tex
NM

The nature of the random fluctuations is revealed by the corresponding histograms shown in Fig.~\ref{ExpDiagStats}(d)--(e) on a semi-log scale. The upstream measurements are symmetrically distributed about the mean, $\eta = 0$. In fact, \boetal found that these measurements follow a Gaussian distribution closely \cite{bolles2019}. The downstream measurements, however, skew strongly towards positive displacement, $\eta > 0$. \boetal found these measurements to be well described by a mean-zero gamma distribution \cite{bolles2019}. The slower decay rate of this downstream distribution indicates an elevated level of extreme surface displacement, i.e.~rogue waves. \boetal estimated that a rogue wave can be up to 65 times more likely in these experiments than if displacements were Gaussian \cite{bolles2019}.

@@

We will only be examining statistics of mean-zero quantities in this paper, and so, for an arbitrary mean-zero quantity $q$, we have the following definitions
\begin{align}
%& \var(q) = \mean{q^2}		&&\mbox{\em variance} \\
& \std(q) = q_{std}= \sqrt{\mean{q^2}}
&&\mbox{\em standard deviation} \\
& \skw(q) = {\mean{q^3}} / {q_{std}^3}	
&&\mbox{\em skewness} \\
& \kurt(q) = {\mean{q^4}} / {q_{std}^4} - 3
&&\mbox{\em (excess) kurtosis}
\end{align}
where $\mean{}$ indicates a mean (here a long-time mean). Hereafter, we will simply refer to the excess kurtosis as kurtosis.

@@

	To complement the displacement statistics, we report here statistics for the surface slope $\eta_x$ in the right column of Fig.~\ref{ExpSpatialStats}. Our attention to slope statistics was motivated by new theoretical developments by \maetal \cite{majda2019} and will be expanded upon in later sections. We extract the surface slope by numerical differentiating images of the free surface using Savitzky-Golay smoothing filters. This ability to extract surface slope is another advantage of our optical measurements over the more commonly used technique of placing a set of discrete wave probes in the tank. As seen in Fig.~\ref{ExpSpatialStats}(a) the standard deviation of slope behaves similar to displacement: $\std(\eta_x)$ increases with driving amplitude, but remains nearly uniform in space for each individual experiment and is not affected by the ADC. Once again though, the higher-order moments respond strongly to the ADC if the driving amplitude is sufficiently high. The skewness of slope becomes highly negative and reaches a minimum near $x = 8$ cm, which is downstream of the ADC but upstream of the position where the displacement statistics peak. The kurtosis of $\eta_x$ reaches a peak at nearly the same location as skewness. We note that the large {\em negative} skewness of $\eta_x$ indicates a bias towards negative slope, which is consistent with a right-moving wave of steep leading surface and shallower trailing surface, i.e.~a wave that is near overturning.

	\boetal systematically explored the effect of varying the driving amplitude on the anomalous statistics for a fixed depth ratio. 
Here, we further explore the effect of driving amplitude, especially in conjunction with theoretical developments, and we also report experiments in which the depth ratio was varied. 
To have a simple baseline for comparison against theory, we refer to the data from Fig.~\ref{ExpDiagStats} as the {\em reference} experiments. In these experiments, the driving amplitude is $\Dphi = 1.38^{\circ}$ (intermediate between the two largest amplitudes shown in Fig.~\ref{ExpSpatialStats}), the upstream depth is $\dup = 12.5$ cm, and the downstream depth is $\ddn = 3$ cm, giving a depth ratio of $\dratdn = \ddn/\dup= 0.24$. From the data represented in Fig.~\ref{ExpSpatialStats}, we extract a characteristic wave amplitude for the representative experiments of $\etastd = 0.21$ cm, which will be important for setting dimensionless parameters that enter the theory. Table \ref{paramtable} lists the range of experimental parameters along with these representative values, as well as values of dimensionless parameters that will be introduced later.
% NOTE: Maybe modify slightly the last paragraph if it does not seem to fit next time I read it.

@@

Now it is possible to choose the scales $\ampscale, \lengthscale, \timescale$ for ease in working with a particular framework. We make the following choices,
\begin{align}
\label{scales}
\ampscale = \pi^{1/2} \, \etastd \, , \qquad
\lengthscale_{\pm} = \frac{\lamfac \lamupdn}{2 \pi} \, , \qquad
\timescale_{\pm} = \frac{\lamfac \lamupdn}{2 \pi \freqp \dupdn}
\end{align}
where $\lamfac$ is an integer to be chosen later. 
The explanation for these choices is as follows. First, we have chosen the characteristic amplitude, $\ampscale$, above to normalize the energy of the state-variable $u$ to unity, as will be demonstrated in Section \ref{tKdVSec}. Second, regarding $\lengthscale$, recall that $\lam$ is the characteristic wavelength corresponding to the peak forcing frequency $\freqp$ in the experiments. If only integer multiples of $\freqp$ were imposed (e.g.~lower frequencies were not present), then the forcing would produce waves that are periodic over lengthscale $\lam$. Since lower frequencies do exist, strict periodicity is not satisfied, but rather waves may be nearly periodic over the physical domain $\xi \in [-\lam/2, \lam/2]$. The approximation of near-periodicity becomes more accurate if integer multiples are considered, i.e.~$\xi \in [-\lamfac \lam/2, \lamfac \lam/2]$. Thus, we have chosen $\lengthscale$ above so that, over the dimensionless domain $\tilde{x} \in [-\pi, \pi]$, periodic boundary conditions can be  imposed on $u$ with an accuracy that increases with $\lamfac$.

	Lastly, regarding the characteristic timescale $\timescale$, the most basic timescale in the experiments is simply $\freqp^{-1}$, i.e.~the period of waves passing a fixed reference point. Of course, the leading-order behavior in shallow water is simple translation of waves with uniform speed $c$. The KdV equation provides the first correction to this behavior and describes dynamics that evolve over longer timescales. Hence we have rescaled $\freqp^{-1}$ by the factor $N \lam/(2 \pi \depth) \gg 1$, which provides a suitable long timescale in line with previous normalizations \cite{johnson1997modern}. The scales $\lengthscale = \lengthscale_{\pm}$ and $\timescale = \timescale_{\pm}$ change value across the ADC, which is important to note when comparing the theory against experimental measurements.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\label{HamStruct}
\partial_t{u} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which will take different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\label{H3H2}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\label{Hamiltonian}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw} gives the separate upstream and downstream Hamiltonians as
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
As seen in \eqref{H3H2}, the cubic component, $\Hthree$, represents the skewness of the wave-field, while the quadratic component, $\Htwo$, represents the energy of the surface slope. The sign difference between the two \eqref{Hamiltonian} thus represents a competition between wave skewness and slope energy. This is the reason for our interest in experimental measurements of the surface {\it slope} statistics in the current manuscript.

@@

We now introduce the truncated KdV (TKdV) system, which is the main focus of this study. Consider the Fourier series of the state variable
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k(t) \, e^{i k x} \, , \\
\label{uhat}
&\uhat_k(t) = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x,t) \, e^{-i k x} \dx
\end{align}
Since $u$ is real valued, $\uhat_{-k} = \uhat_{k}^*$, and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k(t) \, e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Projecting the KdV equation \eqref{dimlessKdV} onto the finite dimensional space gives the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad C_2 = \frac{2 \pi^2 \delup}{3 \lamfac^2}
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. The constants $C_3$ and $C_2$ are the same as before and have been repeated here for convenience.

>>>>

The nature of the random fluctuations is revealed by the corresponding histograms shown in Fig.~\ref{ExpDiagStats}(d)--(e) on a semi-log scale. The upstream measurements are symmetrically distributed about the mean, $\eta = 0$. In fact, \boetal found that these measurements follow a Gaussian distribution closely \cite{bolles2019}. The downstream measurements, however, skew strongly towards positive displacement, $\eta > 0$. \boetal found these measurements to be well described by a mean-zero gamma distribution \cite{bolles2019}. The slower decay of the gamma distribution indicates an elevated level of extreme surface displacement, i.e.~rogue waves. \boetal estimated that a rogue wave can be up to 65 times more likely in these experiments than if displacements were Gaussian \cite{bolles2019}.

@@

In this paper, we only examine statistic of mean-zero quantities, and so, for an arbitrary mean-zero quantity $q$, we have the following definitions
\begin{align}
%& \var(q) = \mean{q^2}		&&\mbox{\em variance} \\
& \std(q) = q_{std}= \sqrt{\mean{q^2}}
&&\mbox{\em standard deviation} \\
& \skw(q) = {\mean{q^3}} / {q_{std}^3}	
&&\mbox{\em skewness} \\
& \kurt(q) = {\mean{q^4}} / {q_{std}^4} - 3
&&\mbox{\em (excess) kurtosis}
\end{align}
where $\mean{}$ indicates a mean --- here a long-time mean at a fixed spatial location. Hereafter, we will simply refer to the excess kurtosis as kurtosis.

@@

	To complement the displacement statistics, we report here statistics for the surface slope $\eta_x$ in the right column of Fig.~\ref{ExpSpatialStats}. Our attention to slope statistics was motivated by new theoretical developments by \maetal \cite{majda2019}, as will be expanded upon in later sections. We extract the surface slope by numerical differentiating images of the free surface using Savitzky-Golay smoothing filters. This ability to extract surface slope is another advantage of our optical measurements over the more commonly used technique of placing a set of discrete wave probes in the tank. As seen in Fig.~\ref{ExpSpatialStats}(a) the standard deviation of slope behaves similar to displacement: $\std(\eta_x)$ increases with driving amplitude, but remains nearly uniform in space for each individual experiment and is not affected by the ADC. The higher-order moments, however, respond strongly to the ADC if the driving amplitude is sufficiently high. The skewness of slope becomes highly negative and reaches a minimum near $x = 8$ cm, which is downstream of the ADC but upstream of the position where the displacement statistics peak. The kurtosis of $\eta_x$ reaches a peak at nearly the same location as skewness. We note that the large {\em negative} skewness of $\eta_x$ indicates a bias towards negative slope, which is consistent with a right-moving wave of steep leading surface and shallower trailing surface, i.e.~a wave that is near overturning.

To have a simple baseline for comparison against theory, we refer to the data from Fig.~\ref{ExpDiagStats} as the {\em reference} experiments. In these experiments, the driving amplitude is $\Dphi = 1.38^{\circ}$ (intermediate between the two largest amplitudes shown in Fig.~\ref{ExpSpatialStats}), the upstream depth is $\dup = 12.5$ cm, and the downstream depth is $\ddn = 3$ cm, giving a depth ratio of $\dratdn = \ddn/\dup= 0.24$. From the data represented in Fig.~\ref{ExpSpatialStats}, we extract a characteristic wave amplitude for the representative experiments of $\etastd = 0.21$ cm, which will be important for setting dimensionless parameters that enter the theory. Table \ref{paramtable} lists the range of experimental parameters and the representative values, as well as values of dimensionless parameters that will be introduced later.

@@

Now it is possible to choose the scales $\ampscale, \lengthscale, \timescale$ for ease in working with a particular framework. We make the following choices,
\begin{align}
\label{scales}
\ampscale = \pi^{1/2} \, \etastd \, , \qquad
\lengthscale_{\pm} = \frac{\lamfac \lamupdn}{2 \pi} \, , \qquad
\timescale_{\pm} = \frac{\lamfac \lamupdn}{2 \pi \freqp \dupdn}
\end{align}
where $\lamfac$ is an integer to be chosen later. 
The explanation for these choices is as follows. First, we have chosen the characteristic amplitude, $\ampscale$, to normalize the energy of the state-variable $u$ to unity, as will be demonstrated in Section \ref{tKdVSec}. Second, regarding $\lengthscale$, recall that $\lam$ is the characteristic wavelength corresponding to the peak forcing frequency $\freqp$ in the experiments. If only integer multiples of $\freqp$ were imposed (e.g.~lower frequencies were not present), then the forcing would produce waves that are periodic over lengthscale $\lam$. Since lower frequencies do exist, strict periodicity is not satisfied, but rather waves may be nearly periodic over the physical domain $\xi \in [-\lam/2, \lam/2]$. The approximation of near-periodicity becomes more accurate if integer multiples are considered, i.e.~$\xi \in [-\lamfac \lam/2, \lamfac \lam/2]$. Thus, we have chosen $\lengthscale$ above so that, over the dimensionless domain $\tilde{x} \in [-\pi, \pi]$, periodic boundary conditions can be  imposed on $u$ with an accuracy that increases with $\lamfac$.

	Lastly, regarding the characteristic timescale $\timescale$, the most basic timescale in the experiments is simply $\freqp^{-1}$, i.e.~the period of waves passing a fixed reference point. Of course, the leading-order behavior in shallow water is simply wave propagation with uniform speed $c$, i.e.~no dispersion. The KdV equation provides the first correction to this behavior and describes dynamics that evolve over longer timescales. Hence we have rescaled $\freqp^{-1}$ by the factor $N \lam/(2 \pi \depth) \gg 1$, which provides a suitably long timescale in line with previous normalizations \cite{johnson1997modern}. The scales $\lengthscale = \lengthscale_{\pm}$ and $\timescale = \timescale_{\pm}$ change value across the ADC, which is important to note when comparing the theory against experimental measurements.

@@

The variable-depth KdV \eqref{dimlessKdV}, though not Hamiltonian throughout the entire domain, admits a Hamiltonian structure on each side of the ADC. Indeed, \eqref{dimlessKdV} can be expressed as
\begin{align}
\label{HamStruct}
\partial_t{u} = \sympJ \vard{\Hupdn}{u}
\end{align}
where $\sympJ = \pdi{}{x}$ is the symplectic operator and $\Ham = \Hupdn$ is the Hamiltonian, which takes different forms on either side of the ADC. It is convenient to decompose the Hamiltonian into a so-called cubic and quadratic component, given respectively by
\begin{align}
\label{H3H2}
\Hthree = \frac{1}{6} \int_{-\pi}^{\pi} u^3 \dx	\, , \qquad
\Htwo = \frac{1}{2} \int_{-\pi}^{\pi} u_x^2 \dx	\, .
\end{align}
Then the Hamiltonian can be expressed as
\begin{equation}
\label{Hamiltonian}
\Hupdn = C_2 \dratupdn^{1/2} \, \Htwo - C_3 \dratupdn^{-3/2} \, \Hthree
\end{equation}
where $\drat = \dratupdn$ changes value across the ADC. More explicitly, substituting \eqref{dratpw} gives the separate upstream and downstream Hamiltonians as
\begin{align}
&\Hup = C_2 \, \Htwo - C_3 \, \Hthree 						&& \text{for } t<0 \\
&\Hdn = C_2 \dratdn^{1/2} \, \Htwo - C_3 \dratdn^{-3/2} \, \Hthree	&& \text{for } t>0
\end{align}
As seen in \eqref{H3H2}, the cubic component, $\Hthree$, represents the skewness of the wave-field, while the quadratic component, $\Htwo$, represents the energy of the surface slope. The sign difference between the two \eqref{Hamiltonian} thus represents a competition between wave skewness and slope energy. In particular, the role played by the slope energy is the motivating factor for the new experimental measurements on surface {\em slope} statistics that are reported in this paper.

@@

We now introduce the truncated KdV (TKdV) system, which is the main focus of the present study. Consider the state variable represented as a spatial Fourier series
\begin{align}
&u(x,t) = \sum_{k=-\infty}^{\infty} \uhat_k(t) \, e^{i k x} \, , \\
\label{uhat}
&\uhat_k(t) = \frac{1}{2 \pi} \int_{-\pi}^{\pi} u(x,t) \, e^{-i k x} \dx \, ,
\end{align}
where $\uhat_k(t) \in \mathbb{C}$. Since $u(x,t)$ is real valued, $\uhat_{-k} = \uhat_{k}^*$, and since momentum vanishes $\uhat_0 = 0$.
Next, consider the Galerkin truncation at wave number $\Lambda$
\begin{align}
\uL(x,t) = \Proj u =
\sum_{\abs{k} \le \Lambda} \uhat_k(t) \, e^{i k x} \, , \qquad
\end{align}
where $\Proj$ is a projection operator and \eqref{uhat} still holds. Inserting the projected variable, $\uL$, into the KdV equation and applying the projection operator, $\Proj$, again where necessary produces the truncated KdV equation (TKdV)
\begin{align}
\label{TKdV}
&\pd{\uL}{t} +  \frac{1}{2} C_3 \drat^{-3/2} \, \pd{}{x} \Proj (\uL)^2 + C_2 \drat^{1/2} \, \frac{\partial^3 \uL}{\partial x^3} = 0
\qquad \text{for } x \in [-\pi,\pi] \\
&C_3 = \frac{3}{2} \pi^{1/2} \epsup \delup^{-1} \, , \quad C_2 = \frac{2 \pi^2 \delup}{3 \lamfac^2}
\end{align}
Note the additional projection operator in front of the quadratic term $\uL^2$, which removes the aliased modes of wavenumber larger than $\Lambda$. Since all wavenumbers larger than $\Lambda$ have been removed, \eqref{TKdV} represents a {\em finite} dimensional dynamical system, of dimension $\Lambda$ over $\mathbb{C}$. The constants $C_3$ and $C_2$ are the same as before and have been repeated here for convenience.
